Reduced growth due to belowground sink limitation is not fully explained by reduced photosynthesis

### 2016-06-15 manuscript/manuscript_pve.Rmd
BM edits, CUE figures

In the absence of a good understanding of the effects of sink limitation on physiological processes, most attempts to model sink limitation of growth to date have done so by abandoning the principle of mass balance (e.g. Leuzinger; CLM nutrient limitation; Fatichi). This approach is unsatisfactory because mechanistic models should obey the laws of physics. Our goal in this paper was to investigate how the carbon balance of plants is modified by sink limitation, by experimentally imposing a sink limitation to growth.

Understanding plant growth and its relationship to carbon (C) assimilation requires knowledge of the mass balance that must be achieved between C uptake and subsequent allocation to growth, storage, and respiration. As woody plants have highly integrated systems of competing carbohydrate sinks [@kozlowski1992carbohydrate], growth is driven by the allocation of photosynthate from source leaves among these different tissue sinks. At long enough time scales leaf photosynthesis (*A*~n~) and respiratory losses together determine net C balance and plant growth. At shorter temporal scales growth can instead be mediated by tissue C storage pools [@chapin1990ecology], which disrupts the coordination between *A*~n~ and growth. This discrepancy has led to the current debate on how strongly plant growth is controlled by either source or sink activity. As plant growth cannot be simply determined by *A*~n~, growth models must take into account several simultaneous processes affecting the C balance at any given time [@fourcaud2008plant]. Whether growth limitation in woody species is driven mostly by *A*~n~ or other components of the C balance remains largely uncertain.

\
In previous studies, the sink limitation of growth has been explored by manipulations of both C source and C sink activity. Examples of experimental approaches that manipulate C source activity include elevated atmospheric CO2 concentration and partial defoliation experiments.

Elevated CO~2~ causes an increase in leaf-level photosynthetic rate, *A*~n~ [@drake1997more; @ainsworth2007response]; across four forest Free-Air CO~2~ Enrichment (FACE) experiments this increased photosynthetic uptake resulted in a stimulation of biomass production by 23 % on average [@norby2005forest]. Evidence from a wide range of elevated CO~2~ experiments, however, also reveals that even with an average photosynthetic enhancement of over 30 %, the biomass growth rate only increases by around 10 % [@kirschbaum2011does].

In partial defoliation experiments, source activity is decreased by reduction of photosynthesizing leaf area. In these experiments, compensatory increases in *A*~n~ of the remaining foliage are commonly found, and have been attributed to several processes including indirect effects on resource availability, such as increaed stomatal conductance [@layne1995end] or enhanced leaf nutrient status [@turnbull2007increased; @pinkard2011gas], or more direct effects of modified sink-source balance such as, reduction in end-product inhibition [@iglesias2002regulation; @zhou2003changes; @handa2005test], enhanced biochemical activity [@ovaska1993b;@layne1995end;@pinkard2011gas], and regulatory sugar signaling [@eyles2013whole]. However, there have been few attempts *(any??)* to determine the overall C balance in these experiments, and it is unclear whether the change in growth following defoliation can be explained by the reduction in leaf area and changes to A~n~

@@

\
Yet another experimental approach is to reduce belowground C sink strength in tree seedlings by manipulating container size and thus rooting volume [@arp1991effects; @nesmith1998effect; @poorter2012pot]. Inadequate rooting volume can decrease C sink strength by progressively restricting root growth in growing plants [@thomas1991root]. In a comprehensive meta-analysis, Poorter et al. [-@poorter2012pot] demonstrated a large effect of container size on plant biomass accumulation. Photosynthetic down-regulation is generally observed in these studies [e.g. @arp1991effects; @mcconnaughay1991physical; @gunderson1994photosynthetic; @sage1994acclimation; @maina2002intra; @ronchi2006growth]. Poorter et al. [-@poorter2012pot] examined the components of relative growth rate and found the unit leaf rate to be most strongly affected. As unit leaf rate is correlated with An, they concluded that down-regulation of An may best explain the effects of container size on biomass growth.

This conclusion arises because plants grown in small containers are shown to accumulate leaf starch while having lower assimilation rates [@robbins1988effect]. However, evidence in support for a trade-off between C storage and growth in trees is, to date, inconclusive [@palacio2014does].  Whether down regulation of *A*~n~ and subsequent reductions of growth are quantitatively similar has rarely been tested. Based on these previous studies, using a range of container sizes as a sink-strength manipulation provdies a way to empirically test the extent to which growth and *A*~n~ are coordinated.

@@

##Growth and morphology
Plant height, diameter and leaf area diverged between container volumes soon after start of the experiment (Figure `r allomfig`a-c). First to diverge was seedling leaf area, which significantly differed among soil volumes (*P* < 0.029) during the 5^th^ week of the experiment. Significant differences were then observed in both height (8^th^ week) and then diameter (9^th^ week) significantly among soil volumes (*P* < 0.045 & 0.035, respectively). The large reductions in height gain and total leaf area in smaller compared to larger containers continued throughout the experiment. In this field study, colder temperatures and reductions in total PPFD per day (Figure `r airvarsfig`) likely slowed the growth of seedlings in the final weeks of the experiment. Seedlings maintained diameter growth throughout the experiment, although growth was marginal in smaller soil volumes in the final month. Final seedling height significantly increased with increasing soil volume (*P* < 0.001). Increases in both final stem diameter and cumulative leaf area (both *P* < 0.001) were found with increasing soil volume. We did not find a continuous effect of container size; rathe, plant size was similar among smaller containers, and these differences were driven mainly by the largest container and the free seedling treatments.

@@

##Modelling seedling biomass
Optimization of the default model (*M*~free~) against data yielded an estimate of LMF of 25.7 %; with this LMF, the model predicted mean harvest total biomass of free seedlings within 1.0 %. Using this optimized LMF, the modeled total biomass of seedlings for each soil volume treatment (*M*~pots~) was on average 11&pm;2.3 g C more than measured seedling biomass when compared against predicted total net leaf C gain (Figure `r massmodelfig`a). As a result, seedling C mass was overestimated by an average of 45&pm;6.2 % in modeled seedlings across the soil volume treatments when non-photosynthetic parameters were kept constant (Figure `r massmodelfig`b). As a result, the observed reductions in leaf *A* with decreasing soil volume when integrated across the 120 day experiment were not large enough to explain the reduction in observed seedling biomass across the container size treatments.

@@

We addressed questions regarding the coordination of *A*~n~ and growth by complementing empirical results with a C mass balance model. We found that reductions in leaf *A*~n~ across container sizes, when integrated across the 120 day experiment, were alone insufficient to account for observed reductions in total plant biomass production.

##Changes in growth and physiology under sink limitation
Soon after seedlings became established, both height and diameter growth were negatively affected by decreasing soil volume. Low growth rates led to large reductions in biomass in small containers, compared to freely rooted seedlings. We analyzed the relationship between biomass growth and soil volume and found an increase of 34 % with a doubling of container volume, consistent with the meta-analysis of Poorter et al. [-@poorter2012pot]. These growth reductions were expected, as the impedance of root growth can cause reductions in overall plant growth and activity [@mcconnaughay1991physical; @young1997mechanical]. It has been shown that roots subjected to environmental stress may send inhibitory signals to the shoots that affect *g*~s~, cell expansion, cell division and the rate of leaf appearance [@passioura2002soil]. Here, we observed a large divergence in leaf area between seedlings in containers and free seedlings through time, with the eventual cessation of new leaf growth in seedlings in small containers.

\
Decreases in *A*~sat~ occurred simultaneously with reductions of growth of seedlings in containers. This observation initially suggests that sink limitation may be driving the down regulation of photosynthesis, and thus growth. However, there are several possible mechanisms that can explain reduced *A*~n~ in small containers including reduced water availability or nutrient content, or reduced sink strength [@poorter2012pot]. It was therefore necessary to examine each of these factors to determine the mechanism by which the induced belowground sink limitation triggered photosynthetic down regulation.

@@

\
The lack of detected shifts in partitioning to fine roots provides evidence against an optimal foraging strategy for seedlings in containers. It is instead possible that lateral root development is affected by inanimate obstacles and avoiding growth towards container walls may improve resource gain [@falik2005root]. The sensitivity of roots to their own exudates near obstructions may also prevent further growth [@semchenko2008foraging]. Here, we found that FRLD was highest in smallest containers suggesting that root restriction likely occurred as simple function of available rooting space. Additionally, physical restriction of root proliferation could have impacted root development and morphology prior to shifts in mass partitioning. Here, increases in SRL were detected in several of the soil volume treatments. This is not surprising as plants in containers have been shown to have different root morphology to field grown plants [@nesmith1998effect]. The poor soil quality used in our experiment and root restriction, however, likely decreased the capacity of this morphological response to increase N uptake.

@@

```{r massmodel, fig.width=6, fig.height=8, echo=FALSE, dev="png", dpi=600, dev.args=list(type="cairo")}
source("master_scripts/massmodel_optimize.R")
```  
**Figure 6**. Total carbon mass for harvested and modeled seedlings versus predicted total carbon gain after 120 days (a) and  reductions in final seedling carbon mass, both modeled and observed, as a function of the reduction in leaf photosynthesis across treatments (b). For (a) the dashed 1:1 identifies the difference between net total leaf carbon gain and gross seedling production. For (b) both seedling carbon mass and daily carbon assimilation were first scaled to the free seedling control.

>>>>

In the absence of a good understanding of the effects of sink limitation on physiological processes, most attempts to model sink limitation of growth to date have done so by abandoning the principle of mass balance *(e.g. Leuzinger; CLM nutrient limitation; Fatichi)*. This approach is unsatisfactory because mechanistic models should obey the laws of physics. Our goal in this paper was to investigate how the carbon balance of plants is modified by sink limitation, by experimentally imposing a sink limitation to growth.

*also deleted below 2 sentences*
As plant growth cannot be simply determined by *A*~n~, growth models must take into account several simultaneous processes affecting the C balance at any given time [@fourcaud2008plant]. Whether growth limitation in woody species is driven mostly by *A*~n~ or other components of the C balance remains largely uncertain.

\
In previous studies, the sink limitation of growth has been explored by manipulations of both C source and C sink activity. Examples of experimental approaches that manipulate C source activity include elevated atmospheric CO2 concentration and partial defoliation experiments. Elevated CO~2~ causes an increase in leaf-level photosynthetic rate, *A*~n~ [@drake1997more; @ainsworth2007response]. While this increase is often observed to drive an increase in plant biomass production (eg any number of papers),

across four forest Free-Air CO~2~ Enrichment (FACE) experiments this increased photosynthetic uptake resulted in a stimulation of biomass production by 23 % on average [@norby2005forest]. Evidence from a wide range of elevated CO~2~ experiments, however, also reveals that even with an average photosynthetic enhancement of over 30 %, the biomass growth rate only increases by around 10 % [@kirschbaum2011does].

\
In partial defoliation experiments, source activity is decreased by reduction of photosynthesizing leaf area. In these experiments, compensatory increases in *A*~n~ of the remaining foliage are commonly found, and have been attributed to several processes including indirect effects on resource availability, such as increaed stomatal conductance [@layne1995end] or enhanced leaf nutrient status [@turnbull2007increased; @pinkard2011gas], or more direct effects of modified sink-source balance such as, reduction in end-product inhibition [@iglesias2002regulation; @zhou2003changes; @handa2005test], enhanced biochemical activity [@ovaska1993b;@layne1995end;@pinkard2011gas], and regulatory sugar signaling [@eyles2013whole]. However, there have been few attempts *(any??)* to determine the overall C balance in these experiments, and it is unclear whether the change in growth following defoliation can be explained by the reduction in leaf area and changes to A~n~

@@

\
Yet another experimental approach is to reduce belowground C sink strength in tree seedlings by manipulating container size and thus rooting volume [@arp1991effects; @nesmith1998effect; @poorter2012pot]. Inadequate rooting volume can decrease C sink strength by progressively restricting root growth in growing plants [@thomas1991root]. In a comprehensive meta-analysis, Poorter et al. [-@poorter2012pot] demonstrated a large effect of container size on plant biomass accumulation. Photosynthetic down-regulation is generally observed in these studies [e.g. @arp1991effects; @mcconnaughay1991physical; @gunderson1994photosynthetic; @sage1994acclimation; @maina2002intra; @ronchi2006growth]. Poorter et al. [-@poorter2012pot] examined the components of relative growth rate and found the unit leaf rate to be most strongly affected. As unit leaf rate is correlated with *A*~n~, they concluded that down-regulation of *A*~n~ may best explain the effects of container size on biomass growth. However, they were unable to determine the physiological mechanism  driving the downregulation of *A*~n~, nor to test whether the reduction in *A*~n~ could explain the observed reduction in growth.

*delete bellow*
This conclusion arises because plants grown in small containers are shown to accumulate leaf starch while having lower assimilation rates [@robbins1988effect]. However, evidence in support for a trade-off between C storage and growth in trees is, to date, inconclusive [@palacio2014does]. Whether down regulation of *A*~n~ and subsequent reductions of growth are quantitatively similar has rarely been tested. Based on these previous studies, using a range of container sizes as a sink-strength manipulation provides a way to empirically test the extent to which growth and *A*~n~ are coordinated.

@@

##Growth and morphology
Plant height, diameter and leaf area diverged between container volumes soon after start of the experiment (Figure `r allomfig`a-c). First to diverge was seedling leaf area, which significantly differed among soil volumes (*P* < 0.029) during the 5^th^ week of the experiment. Significant differences were then observed in both height (8^th^ week) and then diameter (9^th^ week) significantly among soil volumes (*P* < 0.045 & 0.035, respectively). The large reductions in height gain and total leaf area in smaller compared to larger containers continued throughout the experiment. In this field study, colder temperatures and reductions in total PPFD per day (Figure `r airvarsfig`) likely slowed the growth of seedlings in the final weeks of the experiment. Seedlings maintained diameter growth throughout the experiment, although growth was marginal in smaller soil volumes in the final month. Final seedling height significantly increased with increasing soil volume (*P* < 0.001). Increases in both final stem diameter and cumulative leaf area (both *P* < 0.001) were found with increasing soil volume. We did not find a continuous effect of container size; rather, plant size was similar among smaller containers, and differences were driven mainly by the largest container and the free seedling treatments.

@@

##Whole-plant C balance
Optimization of the default model (*M*~free~) against data yielded an estimate of LMF of 25.7 %; with this LMF, the model predicted mean harvest total biomass of free seedlings within 1.0 %. Using this optimized LMF, the modeled total biomass of seedlings for each soil volume treatment (*M*~pots~) was on average 11&pm;2.3 g C more than measured seedling biomass when compared against predicted total net leaf C gain (Figure `r massmodelfig`a). As a result, seedling C mass was overestimated by an average of 45&pm;6.2 % in modeled seedlings across the soil volume treatments when non-photosynthetic parameters were kept constant (Figure `r massmodelfig`b). As a result, the observed reductions in leaf *A* with decreasing soil volume when integrated across the 120 day experiment were not large enough to explain the reduction in observed seedling biomass across the container size treatments.

@@

We addressed questions regarding the coordination of *A*~n~ and growth by complementing empirical results with a C mass balance model.

*Poorter review talks about water availability (rule out) temperature effects on soil volume (can rule out) nutrient availability (possible) root signaling (magic) but strangely, not carbohydrate feedback*

##Changes in growth and physiology under sink limitation
Soon after seedlings became established, both height and diameter growth were negatively affected by decreasing soil volume. Low growth rates led to large reductions in biomass in small containers, compared to freely rooted seedlings. We analyzed the relationship between biomass growth and soil volume and found an increase of 34 % with a doubling of container volume, close to the 43 % reported in the meta-analysis by Poorter et al. [-@poorter2012pot]. These growth reductions were expected, as the impedance of root growth can cause reductions in overall plant growth and activity [@mcconnaughay1991physical; @young1997mechanical]. It has been shown that roots subjected to environmental stress may send inhibitory signals to the shoots that affect *g*~s~, cell expansion, cell division and the rate of leaf appearance [@passioura2002soil]. Here, we observed a large divergence in leaf area between seedlings in containers and free seedlings through time, with the eventual cessation of new leaf growth in seedlings in small containers.

\
Decreases in *A*~sat~ occurred simultaneously with reductions of growth of seedlings in containers. This observation initially suggests that sink limitation may be driving the down regulation of photosynthesis, and thus growth. However, there are several possible mechanisms that can explain reduced *A*~n~ in small containers including reduced water availability or nutrient content, increased soil temperatures, or reduced sink strength [@poorter2012pot]. It was therefore necessary to examine each of these factors to determine the mechanism by which the induced belowground sink limitation triggered photosynthetic down regulation. In this study, the impacts of increasing soil temperatures with decreasing container sizes was controlled by burying of all containers flush to the soil surface.

@@

\
The lack of detected shifts in partitioning to fine roots provides evidence against an optimal foraging strategy for seedlings in containers. It is instead possible that lateral root development is affected by inanimate obstacles and avoiding growth towards container walls may improve resource gain [@falik2005root]. The sensitivity of roots to their own exudates near obstructions may also prevent further growth [@semchenko2008foraging]. Here, we found that FRLD was highest in smallest containers suggesting that root restriction likely occurred as simple function of available rooting space. Additionally, physical restriction of root proliferation could have impacted root development and morphology prior to shifts in mass partitioning. Here, increases in SRL were detected in several of the soil volume treatments. This is not surprising as plants in containers have been shown to have different root morphology to field grown plants [@nesmith1998effect]. The poor soil quality used in our experiment and root restriction, however, likely decreased the capacity of this morphological response to increase N uptake. 
*add exudation*
Alternatively, root exudation may have increased with reduced rooting volumes to facilitate N uptake in favour of increasing partitioning to root biomass.

@@

```{r massmodel, fig.width=6, fig.height=8, echo=FALSE, dev="png", dpi=600, dev.args=list(type="cairo")}
source("master_scripts/massmodel_optimize.R")
```  
**Figure 6**. (a) Total C mass of harvested seedlings versus modelled net leaf carbon gain over 120 days. The dashed black line hows the 1:1 line and the dashed red line shows the seedling biomass that would be expected if all seedlings had the same carbon use efficiency as the freely grown seedling. (b) Carbon use efficiency (ration of final biomass to modelled net leaf C gain) of plants growth in different container sizes.
